My Publication Title --- Single Author

### 2025-07-18 paper/quarto_aaai/paper.tex
Emphasize RQ answers

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}

@@

We seek to answer the following three research questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the CT objective in Equation 1 induce models to
  learn plausible explanations?
\item
  To what extent does CT lead to more favorable AR outcomes in the
  presence of actionability constraints?
\item
  What are the effects of hyperparameter selection on CT?
\end{enumerate}

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). They are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets:
\emph{Adult} (Census data) of \citet{becker1996adult2}, California
housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
(\emph{GMSC}) from \citet{kaggle2011give2}. Finally, for the convenience
of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

@@

Our main results for plausibility and actionability for \emph{MLP}
models are summarised in Table~\ref{tbl-main}, which presents
counterfactual outcomes grouped by dataset along with standard errors
averaged across bootstrap samples.

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. Focusing on the first row for \emph{LS} for
a moment, the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} (panel
(b) vs.~(a)). Asterisks indicate that reductions are statistically
significant: bootstrapped 99\%-confidence intervals of differences in
mean outcomes do \emph{not} include zero.

The third column shows results for the scenario when mutability
constraints are imposed on selected features. Once again we are
comparing CT to the baseline, so reductions in the positive direction
imply that valid counterfactuals are cheaper when using CT with feature
protection. Relating this back to Figure~\ref{fig-poc}, the third column
represents the reduction in distances travelled by counterfactuals in
panel (d) compared to (c). In the following, we summarize the remaining
results.

@@

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Actionability} (column 3): percentage reduction in costs with
protected features. Outcomes are aggregated across bootstrap samples and
varying degrees of the energy penalty \(\lambda_{\text{egy}}\) used for
\emph{ECCo} at test time. Asterisks (\(^*\)) indicate that bootstrapped
99\%-confidence interval of differences in mean outcomes does \emph{not}
include zero.}

@@

\subsubsection{Plausibility.}\label{sec-plaus}

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in
\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
\emph{Circ}; for the real-world tabular datasets they are around 12\%
for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
and generally statistically significant, although the average degree of
uncertainty is higher than for \(\text{IP}\): average reductions range
from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
negative findings for \emph{OL} and \emph{MNIST} are statistically
insignificant and, for MNIST, do not align with qualitative findings,
which are much more plausible for CT (Figure~\ref{fig-mnist}).

@@

\subsubsection{Actionability.}\label{sec-act}

We also find that CT can improve actionability in the presence of
immutable features (column 3 in Table~\ref{tbl-main}): for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five upper and lower pixel rows of the
full image. Reductions in costs are overwhelmingly positive and
significant of up to nearly 60\% for \emph{GMSC}. While the estimated
cost reductions for \emph{Adult} and \emph{MNIST} are not significant,
Figure~\ref{fig-mnist} (columns 6-10) demonstrates that in the latter
case, CT does have the expected effect: sensitivity to protected
features as per integrated gradients is drastically reduced; details of
this experiment are reported in the supplementary appendix. In the case
of \emph{Cred}, average costs increase, likely because any potential
benefits from protecting the \emph{age} are outweighed by the increase
in costs required for greater plausibility.

\subsubsection{Predictive Performance.}\label{sec-pred}

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbations size,
\(\varepsilon \in [0,0.1]\), increases along the horizontal axis and
includes zero, corresponding to standard test accuracy for non-perturbed
data. We observe that for all synthetic datasets, predictive performance
of CT is virtually identical to the baseline and unaffected by
adversarial perturbations. For all real-world datasets, we find that CT
substantially improves robustness: while in some cases baseline
accuracies drop to essentially zero for large enough perturbation sizes,
accuracies for CT are remarkably robust to gradient-based adversarial
attacks.

@@

\subsubsection{Hyperparameter settings.}\label{sec-hyperparameters}

We test the impact of three types of hyperparameters. Here we focus on
the highlights; full results are available in the supplementary
appendix.\\
\indent First, we note that CT is highly sensitive to the choice of a CE
generator and its hyperparameters but (1) there are manageable patterns,
and (2) we can usually identify settings that improve either
plausibility or cost, and often both of them at the same time. For
example, \emph{REVISE} tends to perform the worst, most likely because
it uses a surrogate VAE to generate counterfactuals which impedes
faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the maximum
number of steps, generally yields better outcomes because more CEs can
mature in each training epoch. The impact of \(\tau\), the required
decision threshold is more difficult to predict. On ``harder'' datasets
it may be difficult to satisfy high \(\tau\) for any given sample (i.e.,
also factuals) and so increasing this threshold does not seem to
correlate with better outcomes. In fact, \(\tau=0.5\) generally leads to
optimal results as it is associated with high proportions of mature
counterfactuals.\\
\indent Second, the strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
performance in terms of decreased plausibility and increased costs if
insufficiently high. The sensitivity with respect to
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
evident. While high values of \(\lambda_{\text{reg}}\) may increase the
variability in outcomes when combined with high values of
\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
very pronounced.\\
\indent Third, the effectiveness and stability of CT is positively
associated with the number of counterfactuals generated during each
training epoch. A higher number of training epochs is also beneficial.
Interestingly, we observed desired improvements when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT can improve the
explainability of models in a fine-tuning manner.

@@

We also highlight three key directions for future research. Firstly, it
is an interesting challenge to extend CT beyond classification settings.
Our formulation relies on the distinction between non-target class(es)
and target class(es), requiring the output space to be discrete. Thus,
it does not apply to ML tasks where the change in outcome cannot be
readily discretized. Focus on classification is a common choice in
research on CEs and AR; other settings have attracted some interest,
e.g., regression \citep{spooner2021counterfactual}, but there is little
consensus how to robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities, as also
recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
to two sources of instabilities: (1) the energy-based contrastive
divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
(2) the underlying explainers. We find several promising ways to
mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
generating sufficiently many counterfactuals during each epoch, and
including only mature counterfactuals in \(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future work could benefit from
more sophisticated approaches.

>>>>

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\usepackage{enumitem}

@@

We seek to answer the following four research questions:

\begin{enumerate}[label={(\makebox[2em][c]{RQ\arabic*})}, leftmargin=3.5em]
    \item To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
    \item To what extent does CT result in more favorable algorithmic recourse outcomes in the presence of actionability constraints?
    \item To what extent does CT influence the adversarial robustness of trained models?
    \item What are the effects of hyperparameter selection on counterfactual training?
\end{enumerate}

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). They are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets:
\emph{Adult} (Census data) of \citet{becker1996adult2}, California
housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
(\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the convenience
of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

@@

Our main results for plausibility and actionability for \emph{MLP}
models are summarised in Table~\ref{tbl-main}, which presents
counterfactual outcomes grouped by dataset along with standard errors
averaged across bootstrap samples. Asterisks (\(^*\)) indicate that the
observed effects are statistically significant: bootstrapped
99\%-confidence intervals of differences in mean outcomes do \emph{not}
include zero.

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. As an example, consider the first row
(\emph{LS}): the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} in panel
(b) compared to panel (a).

The third column shows the results for a scenario when mutability
constraints are imposed on the selected features. Again, we are
comparing CT to the baseline, so reductions in the positive direction
imply that valid counterfactuals are ``cheaper'' (more actionable) when
using CT with feature protection. Relating this back to
Figure~\ref{fig-poc}, the third column represents the reduction in
distances travelled by counterfactuals in panel (d) compared to panel
(c). In the following paragraphs, we summarize the results for all
datasets.

@@

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
in costs when selected features are protected. Outcomes are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
differences in mean outcomes does \emph{not} include zero.}

@@

\subsubsection{Plausibility (RQ1).}\label{sec-plaus}

\emph{CT generally produces substantial and statistically significant
improvements in plausibility.}\\
Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings for \emph{OL} and
\emph{MNIST} are not statistically significant and, for MNIST, do not
align with qualitative findings, which are much more plausible for CT
(Figure~\ref{fig-mnist}).

@@

\subsubsection{Actionability (RQ2).}\label{sec-act}

\emph{CT tends to improve actionability in the presence of immutable
features, but the exact effects can be challenging to establish due to
feature interactions.}

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Reductions in costs are overwhelmingly positive
and significant of up to around 60\% for \emph{GMSC}. While the
estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
are not significant and Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that in the latter case, CT does have the expected effect:
sensitivity to protected features as per integrated gradients is
drastically reduced; details of this experiment are reported in the
supplementary appendix. In the case of \emph{Cred}, average costs
increase, likely because any potential benefits from protecting the
\emph{age} are outweighed by the increase in costs required for greater
plausibility.

@@

\subsubsection{Predictive Performance (RQ3).}\label{sec-pred}

\emph{Models trained with CT are substantially more robust to
gradient-based adversarial attacks than conventionally-trained
baselines.}

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
increases along the horizontal axis and includes zero, corresponding to
standard test accuracy for non-perturbed data. For all synthetic
datasets, predictive performance of CT is virtually identical to the
baseline and unaffected by perturbations. For all real-world datasets,
we find that CT substantially improves robustness: while in some cases
baseline accuracies drop to essentially zero for large enough
perturbation sizes, accuracies of CT models are remarkably robust to
gradient-based adversarial attacks.

\subsubsection{Hyperparameter settings
(RQ4).}\label{sec-hyperparameters}

\emph{CT is highly sensitive to the choice of a CE generator and its
hyperparameters but (1) we observe manageable patterns, and (2) we can
usually identify settings that improve either plausibility or
actionability, and typically both of them at the same time.}

We evaluate the impacts of three types of hyperparameters on CT. In this
section we focus on the highlights and make the full results available
in the supplementary appendix.

For example, \emph{REVISE} tends to perform the worst, most likely
because it uses a surrogate VAE to generate counterfactuals which
impedes faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the
maximum number of steps, generally yields better outcomes because more
CEs can mature in each training epoch. The impact of \(\tau\), the
required decision threshold is more difficult to predict. On ``harder''
datasets it may be difficult to satisfy high \(\tau\) for any given
sample (i.e., also factuals) and so increasing this threshold does not
seem to correlate with better outcomes. In fact, \(\tau=0.5\) generally
leads to optimal results as it is associated with high proportions of
mature counterfactuals.

Second, the strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
performance in terms of decreased plausibility and increased costs if
insufficiently high. The sensitivity with respect to
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
evident. While high values of \(\lambda_{\text{reg}}\) may increase the
variability in outcomes when combined with high values of
\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
particularly pronounced.

Third, the effectiveness and stability of CT is positively associated
with the number of counterfactuals generated during each training epoch.
A higher number of training epochs is also beneficial. Interestingly, we
observed desired improvements when CT was combined with conventional
training and applied only for the final 50\% of epochs of the complete
training process. Put differently, CT can improve the explainability of
models in a fine-tuning manner.

@@

We also highlight three key directions for future research. Firstly, it
is an interesting challenge to extend CT beyond classification settings.
Our formulation relies on the distinction between non-target class(es)
and target class(es), requiring the output space to be discrete. Thus,
it does not apply to ML tasks where the change in outcome cannot be
readily discretized. Focus on classification is a common choice in
research on CEs and AR; other settings have attracted some interest,
e.g., regression \citep{spooner2021counterfactual}, but there is little
consensus how to robustly extend the notion of CEs.

Secondly, CT is susceptible to training instabilities, as is also the
case for the related JEMs \citep{grathwohl2020your}. CT is exposed to
two sources of instabilities: (1) the energy-based contrastive
divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
(2) the underlying explainers. We find several promising ways to
mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
generating sufficiently many counterfactuals during each epoch, and
including only mature counterfactuals in \(\text{div}(\cdot)\).

Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future work could benefit from
more sophisticated approaches.
